---
title: "Statistical Rethinking 23 - Homework"
author: "Bo Bindzus Jespersen (DKBOBJ)"
date: '2023-02-15'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(devtools)
library("rethinking")
rstan_options(auto_write = TRUE) # To avoid recompilation of unchanged Stan programs
library("ggdag")
library(ggplot2)
```

# Lecture 02: Garden of Forking Data

1.  Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out
    to be 4 water and 11 land. Construct the posterior distribution.

```{r}
sample <- c(rep("W",4),rep("L",11))
W <- sum(sample=="W")
L <- sum(sample=="L")
p <- seq(0,1,0.25)
ways <- sapply(p, function(q) (4*q)^W * ((1-q)*4)^L)
prob <- ways/sum(ways)
cbind(p, ways, prob)


curve(dbeta(x, 4+1, 11+1), 
      xlab = "proportion of water",
      ylab = "density"
      )

```

2.  Using the posterior distribution from 1, compute the posterior
    predictive distribution for the next 5 tosses of the same globe. I
    recommend you use the sampling method.

```{r}
n <- 10000
p_sample <- rbeta(n, 4+1, 11+1)
w_sim <- rbinom(n, 5, p_sample) # rbinom (# observations, # trails/observation, probability of success )
# a thousand times we repeat a trial 5 times with a probability of success of p_sample (proportion of water).
# the number represents the number of globe tosses that results in water for a given sampled proportion of water (p_water)
plot(table(w_sim)/sum(table(w_sim)))
```

3.  Use the posterior predictive distribution from 2 to calculate the
    probability of 3 or more water samples in the next 5 tosses.

```{r}
sum(w_sim>=3)/n
```

4-OPTIONAL. This problem is an optional challenge for people who are
taking the course for a second or third time. Suppose you observe W = 5
water points, but you forgot to write down how many times the globe was
tossed, so you don't know the number of land points L. Assume that p =
0.7 and compute the posterior distribution of the number of tosses N.
Hint: Use the binomial distribution.

# Lecture 03 & 04: Geocentric Models & Categories and Curves

1.  From the Howell1 dataset, consider only the people younger than 13
    years old. Estimate the causal association between age and weight.
    Assume that age influences weight through two paths. First, age
    influences height, and height influences weight. Second, age
    directly influences weight through age-related changes in muscle
    growth and body proportions. Draw the DAG that represents these
    causal relationships. And then write a generative simulation that
    takes age as an input and simulates height and weight, obeying the
    relationships in the DAG.
    
```{r fig.height=3}
# Age influence weight through height
dagify(Height ~ Age,
       Weight ~ Height + Age,
       Weight ~ U,
       Height ~ U) |> 
  ggdag() +
  theme_void()
  
```

```{r}
# Simulating both the direct and indirect influences of weight with uncertanties for both. 

sim_W <- function(A, bAH=5, bHW=0.5, bAW=0.5, sdU=2){
  n <- length(A)
  # add rnorm to both H and W
  # First the indirect influence of H is calculated incl. some uncertainty (rnorm)
  H = bAH*A+rnorm(n, sd=sdU)
  # Then the direct influence on W is calculated inlc. some new uncertainty.
  W = bHW*H + bAW*A + rnorm(n, sd=sdU)
  return(data.frame(age=A, height=H, weight=W))
}

sim_W(runif(20,1,12)) |> 
  pairs()
```



2.  Use a linear regression to estimate the total causal effect of each
    year of growth on weight.

```{r}
A <- seq(1,20,0.5)
sim_W(A=A, bAH = 0.5, bHW = 0.5, bAW = 0.5, sdU=5)

data(Howell1)
h <- Howell1[Howell1$age<13,]  # Only kids yonger than 13 are included


sim1 <- sim_W(A=h$age) # simulating weight based on height from dataset (based on priors)

with(sim1, plot(age, height))

# Quadratic approximation of height on weight
m1 <- quap(
  alist(
    # The order is EXTREMELY important!!!!
    W ~ dnorm(mu, sigma),
    mu <- a + b*A,
    a ~ dnorm(0, 10),
    b ~ dunif(0,5),
    sigma ~ dexp(1) #????? why exp???
  ),
  data = list(
    A=h$age,
    W=h$weight 
    )
)

# b is the casual relationship
precis(m1)
```


3. Now suppose the causal association between age and weight might be
different for boys and girls. Use a single linear regression, with a
categorical variable for sex, to estimate the total causal effect of age
on weight separately for boys and girls. How do girls and boys differ?
Provide one or more posterior contrasts as a summary.

```{r fig.height=3}
# Age influence weight through height
dagify(Height ~ Age,
       Weight ~ Height + Age,
       Weight ~ Sex,
       Height ~ Sex) |> 
  ggdag() +
  theme_void()
  
```




```{r}
m2 <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a[S] + b[S]*A,
    a[S] ~ dnorm(0, 10),
    b[S] ~ dunif(1,5),
    sigma ~ dexp(1)
  ),
  data = list(
    A = h$age,
    W = h$weight,
    S = h$male+1
  )
)

precis(m2, depth=2)

# The posterior of the intercept (a) for boys and girls:
post <- extract.samples(m2)

# Drawing random variables from the posterior 
post1 <- rnorm(1000, post$a[,1], post$sigma)
post2 <- rnorm(1000, post$a[,2], post$sigma)

# plotting posteriors
dens(post1, col=2, main="Posterior for the intercept")
dens(post2, add = T, col=4)
```




```{r}
age <- 0:13

# link is used to compute values of any linear models over samples from the posterior distribution. 
link1 <- link(m2, data=list(A=age, S=rep(1,14)))
link2 <- link(m2, data=list(A=age, S=rep(2,14)))

plot( NULL , xlim=c(0,13) , ylim=c(5,25), xlab="age", ylab="weight")
points(x=h$age[h$sex==1],y=h$weight[h$sex==1], col=2)
points(x=h$age[h$sex==2],y=h$weight[h$sex==2], col=4)
shade(apply(link1,2,PI,prob=0.89), age)
shade(apply(link2,2,PI,prob=0.89), age)
abline(a = 7.04, b = 1.29, col=2)
abline(a = 7.87, b = 1.39, col=4)
```

Plotting the contrast difference of weight between boys and girls.

```{r}
# age interval for simulating "people" using the model
age <- 0:13

# sim is used to simulate posterior predictive distributions, simulating outcomes over samples from the posterior distribution of parameters.
# Simulating the weight. The Average weight can be simulated using link function
girlsSim <- sim(m2, data=list(A=age, S=rep(1,14)))
boysSim <- sim(m2, data=list(A=age, S=rep(2,14)))

contrast <- boysSim-girlsSim

plot( NULL , xlim=c(0,13) , ylim=c(-15,15) , xlab="age" ,
ylab="weight difference (boys-girls)" )
for ( p in c(0.5,0.67,0.89,0.99) )
shade( apply(contrast,2,PI,prob=p) , age )
abline(h=0, lty=2)
abline(a = 7.87-7.04, b = 1.39-1.29)
```




# Lecture 05: Elemental Confounds
# Lecture 06: Good and Bad Controls
# Lecture 07: Overfitting
# Lecture 08: MCMC
# Lecture 09: Modeling Events
# Lecture 10: 
# Lecture 11: 
# Lecture 12: 
# Lecture 13: 
# Lecture 14: 
# Lecture 15: 




# Week 03:  / 

# Week 04:  / 

# Week 05:  / Counts and Confounds

# Week 06: Ordered Categories / Multilevel Models

# Week 07: Multilevel Adventures / More Multilevel Models

# Week 08: Social networks / Gaussian Processes

# Week 09: Measurement Error / Missing Data

# Week 10: Beyond GLMs: State-space Models, ODEs / Horoscopes
